### 2022-02-11 chapters/conclusions/conclusions.tex
added figures to intro paragraph about hydro

\subsection*{Better understanding the protective role of corals against storm waves}
Simulation results from the coupled wave-current model suggested that storm waves were significantly dissipating over reefs. The modeled significant wave height experiences a twofold reduction over the barrier reef during the passage of the hurricane (Fig. \ref{ccl:swh}). This highlights the important protective role of coral reefs against coastal flood hazard. Better understanding this role can motivate coral reef restoration efforts. For instance, \citep{ferrario2014effectiveness} estimated that coral reefs could provide wave attenuation comparable to artificial defenses such as breakwaters for a lower cost. However, modelling the wave-flow dynamics at the reef scale is particularly challenging and requires models that can explicitly represent small-scale nonhydrostatic processes. The smallest scales that model such as SLIM and SWAN can simulate is limited by the intrinsic assumptions in the model formulations, i.e., hydrostatic for SLIM and phase-averaged for SWAN. These assumptions are no longer valid as soon as the horizontal scales of motion become comparable to the vertical scales \citep{marshall1997hydrostatic}. For such situations, it becomes necessary to couple the SLIM+SWAN model with a phase-resolving nonhydrostatic wave-flow model, hence achieving a multi-physics modelling framework. Nonhydrostatic models are more expensive to run as they require a much finer spatial resolution (~1 m) and have a more complex physics. They can therefore often be used only in rather small areas and over short periods of time \citep{fringer2019future}. One promising approach to tackle this issue would be to use machine learning algorithms to locally emulate the dynamics of a complex high-resolution model and hence speed up the computational process \citep{kasim2021building}. Machine learning is based on the principle of extracting information from data, which can either be observational data or high-resolution model outputs, and deriving relationships between datasets without the need for physics-based models. This procedure has recently been applied to ocean modelling and was able to reproduce the temporal and spatial variability of unresolved turbulent processes, when given a smoothed view of the dynamics \citep{bolton2019applications}.

>>>>

\subsection*{Better understanding the protective role of corals against storm waves}
Simulation results from the coupled wave-current model suggested that storm waves were significantly dissipating over reefs. The modeled significant wave height experiences a twofold reduction over the barrier reef during the passage of the hurricane (Fig. \ref{ccl:swh}). This highlights the important protective role of coral reefs against coastal flood hazard. Better understanding this role can motivate coral reef restoration efforts. For instance, \citep{ferrario2014effectiveness} estimated that coral reefs could provide wave attenuation comparable to artificial defenses such as breakwaters for a lower cost. However, modelling the wave-flow dynamics at the reef scale is particularly challenging and requires models that can explicitly represent small-scale nonhydrostatic processes. The smallest scales that model such as SLIM and SWAN can simulate is limited by the intrinsic assumptions in the model formulations, i.e., hydrostatic for SLIM and phase-averaged for SWAN. These assumptions are no longer valid as soon as the horizontal scales of motion become comparable to the vertical scales \citep{marshall1997hydrostatic}. For such situations, it becomes necessary to couple the SLIM+SWAN model with a phase-resolving nonhydrostatic wave-flow model, hence achieving a multi-physics modelling framework. Nonhydrostatic models are more expensive to run as they require a much finer spatial resolution ($\sim$ 1 m) and have a more complex physics. They can therefore often be used only in rather small areas and over short periods of time \citep{fringer2019future}. One promising approach to tackle this issue would be to use machine learning algorithms to locally emulate the dynamics of a complex high-resolution model and hence speed up the computational process \citep{kasim2021building}. Machine learning is based on the principle of extracting information from data, which can either be observational data or high-resolution model outputs, and deriving relationships between datasets without the need for physics-based models. This procedure has recently been applied to ocean modelling and was able to reproduce the temporal and spatial variability of unresolved turbulent processes, when given a smoothed view of the dynamics \citep{bolton2019applications}. Additionally, this multi-physics modeling frameworks could be uses to better capture the motion of coral larvae over reef. This would improve the modeled coral connectivity, as the the initial motion of larvae over reefs then drives larval transport following mass spawning events and initiate coral connectivity.

### 2022-02-11 chapters/intro/intro.tex
added figures to intro paragraph about hydro

The Florida Reef Tract (FRT) spans over approximately 577 km from the Dry Tortugas (DRTO) west of the Florida Keys to the St. Lucie Inlet in Martin County, constituting the third largest barrier reef in the world \citep{finkl2008shelf}. These reefs harbor a rich Caribbean fauna, including more than 40 species of stony corals \citep{banks2008reef,jackson2014status}. The northern half-section of the FRT consists of relic, Holocene framework reefs and indurated sand ridges, divided into three shore-parallel reefs (inner, middle, outer) separated by sandy plains. Benthic cover is generally denser on the middle and outer reefs while the inner reef harbors some large patches of dense \textit{Acropora cervicornis}. The dominant reef builder is the bouldering \textit{Montastraea cavernosa} but living hard coral cover is fairly low (below 3\%)\citep{banks2008reef,walton2018impacts}. The southern half-section is a chain of limestone islands (the Keys) that extend from the southern tip of the Florida mainland southwest to the DRTO. The Upper Keys are remnants of ancient coral reefs and the Lower Keys are sand bars \citep{hoffmeister1968geology}. The DRTO are made of small circular reef banks whose coral populations are among the most preserved of the continental United States \citep{hine2008coral, kourafalou2018physical}. Moreover, they are believed to be important sources of recruitment for coral-reef fishes in the Florida Keys \citep{domeier2004potential}.

\todo{Add figure with map of the FRT}

@@

\todo{Add map of the propagation of SCTLD}

@@

The latest major outbreak affecting coral reefs of the Caribbean is the SCTLD \citep{noaa2018}. First observed off the coasts of Miami in 2014 during the PMDDP \citep{precht2016unprecedented}, the disease has since spread throughout the entire Florida Reef Tract and numerous territories of the Caribbean \citep{alvarez2019rapid, kramer2019map, estrada2021effects}. The outbreak affects at least 24 species of scleractinian coral and often results in whole colony mortality \citep{precht2016unprecedented, walton2018impacts}. Briefly, the gross morphology of SCTLD can be focal or multifocal, with locally extensive to diffuse areas of acute to subacute tissue loss distributed basally, peripherally, or both. In some cases, tissues bordering areas of chronic tissue loss show indistinct bands (1–5 cm) of pallor, progressing to normal pigmentation away from the denuded skeleton (Fig. \ref{intro:sctld}). The continued persistence of the outbreak, the high number of species affected, and its large geographical distribution suggest that SCTLD might be the largest coral disease outbreak on record in Florida. Although the causative agent of the disease remains unknown, there is evidence of waterborne disease transmission \citep{aeby2019pathogenesis,eaton2021measuring,meiling2021variable} and recent studies showed evidence that sediments can act as a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.

@@

Additionally, Florida is a prime landfall target for tropical cyclones from June through November (Fig. \ref{inro:landfall}). Between 1899 and 1998, Southeastern Florida has been directly hit by 27 hurricanes, 5 of which were of category 4 or more (winds $>$ 209 km/h) \citep{neumann1999tropical}. Hurricanes that forms in June and July are usually weak while hurricanes occurring in August and September tend to become severe storms \citep{banks2008reef}. This is illustrated by Hurricane Irma, one of the strongest and costliest hurricanes on record in the Atlantic, that made landfall in Florida in September 2017 as a category 4 hurricane \citep{irmaNOAA, xian2018brief}. Hurricanes are a major agent of coral mortality in the Caribbean that contributed to the decline of acroporids \citep{gardner2003long,aronson2001white}. They cause a wholesale destruction of the reef biota on their passage and can induce coral burial through sediment resuspension \citep{banks2008reef, miller2008effects}. Furthermore, they significantly impact the content of surface waters through upwelling and mixing, which can disturb the benthic communities \citep{wachnicka2019hurricane,varlas2020investigating}. Moreover, heavy wind conditions generate important waves that interact with currents and significantly alter transport processes in nearshore waters \citep{niu2017role,mao2020particle}. This might impact the dispersion of coral larvae, that are released during the hurricane season \citep{frys2020fine}. Hurricane might therefore interrupt larval exchanges between otherwise connected reef but also create new connectivity pathways. As major hurricanes are intensifying under the effect global warming \citep{bhatia2019recent,knutson2020tropical}, better understanding the impact of hurricane on wave-induced processes could inform the management of reefs under future climatic conditions. Furthermore, as incoming storm-generated waves dissipate over reefs, better understanding the protective role of coral reefs might promote their restoration as tool to reduce coastal flooding hazard \citep{roelvink2021coral}.

@@

\section*{Modeling the hydrodynamics of the FRT}
\todo{Description of circulation in the FRT and SLIM ?}

\section*{Outline}
This thesis is mainly a collection of papers describing our answers to the four main research questions. Chapter 2 describes the development of a coupled epidemio-hydrodynamic model to reproduce the observed spread of SCTLD in the FRT. Chapter 3 explains how this model was used to study the apparent stalling of the spread of the SCTLD between the Marquesas and the DRTO. Chapter 4 investigates the impact of the PMDDP on the onset of the SCTLD oubreak in 2014. Finally, chapter describes the development of a coupled wave-current model to study the impacts of Hurricane Irma (2017) on wave-induced transport processes in the Florida Keys.

@@

    \item Holstein, D. M., \textbf{Dobbelaere, T.}, Gramer, L. J., McEachron, L., Muller E. M., Williams, S. D. \&S Hanert, E. (2022) Coral restoration in the age of emergent disease: Balancing the benefits and risks of population connectivity. \textit{Conservation Letters}. in preparation.

@@

\begingroup
    \renewcommand{\arraystretch}{1.3}
    \begin{tabular}{ll}
        \textbf{DRTO}  & Dry Tortugas \\
        \textbf{ECMWF} &  \\
        \textbf{FCR}   & Florida's Coral Reef \\
        \textbf{FRT}   & Florida Reef Tract  \\
        \textbf{HYCOM} & \\
        \textbf{NOAA}  & National Oceanographic and Atmospheric Administration \\
        \textbf{PoM}   & Port of Miami \\
        \textbf{PMDDP} & PortMiami Deep Dredge Project \\
        \textbf{RS}    & Radiation stress \\
        \textbf{SCTLD} & Stony coral tissue loss disease \\
        \textbf{SLIM}  & \\
        \textbf{SWAN}  &
    \end{tabular}
\endgroup

>>>>

The Florida Reef Tract (FRT) spans over approximately 577 km from the Dry Tortugas (DRTO) west of the Florida Keys to the St. Lucie Inlet in Martin County, constituting the third largest barrier reef in the world \citep{finkl2008shelf} (Fig. \ref{intro:frt}). These reefs harbor a rich Caribbean fauna, including more than 40 species of stony corals \citep{banks2008reef,jackson2014status}. The northern half-section of the FRT consists of relic, Holocene framework reefs and indurated sand ridges, divided into three shore-parallel reefs (inner, middle, outer) separated by sandy plains. Benthic cover is generally denser on the middle and outer reefs while the inner reef harbors some large patches of dense \textit{Acropora cervicornis}. The dominant reef builder is the bouldering \textit{Montastraea cavernosa} but living hard coral cover is fairly low (below 3\%)\citep{banks2008reef,walton2018impacts}. The southern half-section is a chain of limestone islands (the Keys) that extend from the southern tip of the Florida mainland southwest to the DRTO on the the boundary of the West Florida Shelf (WFS). The Upper Keys are remnants of ancient coral reefs and the Lower Keys are sand bars \citep{hoffmeister1968geology}. The DRTO are made of small circular reef banks whose coral populations are among the most preserved of the continental United States \citep{hine2008coral, kourafalou2018physical}. Moreover, they are believed to be important sources of recruitment for coral-reef fishes in the Florida Keys \citep{domeier2004potential}.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/intro/figures/frt.png}
    \caption{Map of Florida Reef Tract with major features and divisions. Source: \cite{kupfner2019untapped}}
    \label{intro:frt}
\end{figure}

@@

@@

The latest major outbreak affecting coral reefs of the Caribbean is the SCTLD \citep{noaa2018}. First observed off the coasts of Miami in 2014 during the PMDDP \citep{precht2016unprecedented}, the disease has since spread throughout the entire Florida Reef Tract and numerous territories of the Caribbean \citep{alvarez2019rapid, kramer2019map, estrada2021effects} (Fig. \ref{intro:propagation}). The outbreak affects at least 24 species of scleractinian coral and often results in whole colony mortality \citep{precht2016unprecedented, walton2018impacts}. Briefly, the gross morphology of SCTLD can be focal or multifocal, with locally extensive to diffuse areas of acute to subacute tissue loss distributed basally, peripherally, or both. In some cases, tissues bordering areas of chronic tissue loss show indistinct bands (1–5 cm) of pallor, progressing to normal pigmentation away from the denuded skeleton (Fig. \ref{intro:sctld}). The continued persistence of the outbreak, the high number of species affected, and its large geographical distribution suggest that SCTLD might be the largest coral disease outbreak on record in Florida. Although the causative agent of the disease remains unknown, there is evidence of waterborne disease transmission \citep{aeby2019pathogenesis,eaton2021measuring,meiling2021variable} and recent studies showed evidence that sediments can act as a vector for the SCTLD \citep{rosales2020rhodobacterales, studivan2022reef}.

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{chapters/intro/figures/fig_sctld.png}
    \caption{Propagation of the SCTLD troughout the FRT between 2014 and 2021}
    \label{intro:propagation}
\end{figure}

@@

Additionally, Florida is a prime landfall target for tropical cyclones from June through November (Fig. \ref{inro:landfall}). Between 1899 and 1998, Southeastern Florida has been directly hit by 27 hurricanes, 5 of which were of category 4 or more (winds $>$ 209 km/h) \citep{neumann1999tropical}. Hurricanes that forms in June and July are usually weak while hurricanes occurring in August and September tend to become severe storms \citep{banks2008reef}. This is illustrated by Hurricane Irma, one of the strongest and costliest hurricanes on record in the Atlantic, that made landfall in Florida in September 2017 as a category 4 hurricane \citep{irmaNOAA, xian2018brief}. Hurricanes are a major agent of coral mortality in the Caribbean that contributed to the decline of acroporids \citep{gardner2003long,aronson2001white}. They cause a wholesale destruction of the reef biota on their passage and can induce coral burial through sediment resuspension \citep{banks2008reef, miller2008effects, tweel2014contribution}. Furthermore, they significantly impact the content of surface waters through upwelling and mixing, which can disturb the benthic communities \citep{wachnicka2019hurricane,varlas2020investigating}. Moreover, heavy wind conditions generate important waves that interact with currents and significantly alter transport processes in nearshore waters \citep{niu2017role,mao2020particle}. This might impact the dispersion of coral larvae, that are released during the hurricane season \citep{frys2020fine}. Hurricane might therefore interrupt larval exchanges between otherwise connected reef but also create new connectivity pathways. As major hurricanes are intensifying under the effect global warming \citep{bhatia2019recent,knutson2020tropical}, better understanding the impact of hurricane on wave-induced processes could inform the management of reefs under future climatic conditions. Furthermore, as incoming storm-generated waves dissipate over reefs, better understanding the protective role of coral reefs might promote their restoration as tool to reduce coastal flooding hazard \citep{roelvink2021coral}.

@@

\section{Modeling the hydrodynamics of the FRT}
The FRT is located along the northern side of the Straits of Florida, that connects the Gulf of Mexico (GoM) and the North Atlantic Ocean. The large-scale circulation in this region is dominated by the interplay of the Florida Current (FC) and the Loop Current (LC). The LC is an area of warm waters from the Caribbean Sea that enters the GoM through the Yucatan Channel and then loops anti-cyclonically through the northeastern Gulf of Mexico and finally exits through the southern Straits of Florida, where it forms the Florida Current (FC). The FC then flows northward between Florida and the Great Bahama Bank to form the Gulf Stream. The northward penetration of the LC through the GoM varies throughout the year, and sporadically an anticyclonic ring separates from the current \citep{leipper1970sequence, maul1977annual, vukovich1988loop}. The detachment of these anticyclonic eddies from the main LC affects in turn the position of the FC, causing its meandering along the FRT. These variations of the FC position have been shown to have no seasonal pattern with strong interannual variability. The generated meandering is associated with the presence of cyclonic eddies between the core of the current and the FRT \citep{kourafalou2012florida}. These eddies significantly impact the ecology of the southwest Florida Shelf as they have been identified as a mechanism causing larval retention on the shelf and delivering larval pulses to the Upper Keys \citep{lee1994evolution, sponaugle2005florida, kourafalou2012florida}. Furthermore, when the LC meets the shallow waters of the southwest corner of the WFS near the DRTO, it causes the inflow of nutrients that modify water properties on the shelf \citep{weisberg2003local, liu2016offshore}. These contacts have been hypothesized to impact in turn the penetration of the LC in the GoM \citep{weisberg2017loop}. On the northernmost section of the FRT however, eddies generated by the meandering of the FC cause cold-water upwelling limiting the northern range of tropical species \citep{walker2013determining}. The circulation on the upper part of the shelf along the FRT is largely influenced by winds, as well as diurnal and semi-diurnal tidal processes \citep{lee2001transport, lee2002volume,d2007patterns}

\todo{Figure LC, GoM Florida Straits, etc.}

The aforementioned large-scale ocean circulation significantly impact transport processes in the FRT. The modeling tools developed in this thesis should therefore be able to capture them accurately. Furthermore, as our objective is to model the dispersal of disease agents and sediments in a topologically complex coastal reef ecosystem, our modeling framework should also be able to capture small-scale ocean circulation. For instance, small-scale flow features such as recirculation eddies around reefs and islands have been shown to strongly influence larval dispersal by increasing their local retention over the reefs \citep{figueiredo2013synthesizing}. Furthermore, accurately capturing the topography of coastal areas is critical when modeling storm conditions, as large wind waves generate strong currents, waves and storm surges in nearshore regions \citep{dietrich2010high, weisberg2006hurricane}. Unstructured-mesh models offer a well-suited solution to tackle the multi-scale nature of flow features in the FRT.

\section*{Outline}
This thesis is mainly a collection of papers describing our answers to the four main research questions. Chapter \ref{chap:sctld} describes the development of a coupled epidemio-hydrodynamic model to reproduce the observed spread of SCTLD in the FRT. Chapter \ref{chap:drto} explains how this model was used to study the apparent stalling of the spread of the SCTLD between the Marquesas and the DRTO. Chapter \ref{chap:onset} investigates the impact of the PMDDP on the onset of the SCTLD oubreak in 2014. Finally, chapter \ref{chap:irma} describes the development of a coupled wave-current model to study the impacts of Hurricane Irma (2017) on wave-induced transport processes in the Florida Keys.

@@

    \item Holstein, D. M., \textbf{Dobbelaere, T.}, Gramer, L. J., McEachron, L., Muller E. M., Williams, S. D. \& Hanert, E. (2022) Coral restoration in the age of emergent disease: Balancing the benefits and risks of population connectivity. \textit{Conservation Letters}. in preparation.

@@

\begingroup
    \setlength{\tabcolsep}{20pt}
    \renewcommand{\arraystretch}{1.3}
    \begin{tabular}{ll}
        \textbf{DRTO}  & Dry Tortugas \\
        \textbf{ECMWF} & European Centre for Medium-Range Weather Forecast \\
        \textbf{FC}    & Florida Current \\
        \textbf{FCR}   & Florida's Coral Reef \\
        \textbf{FRT}   & Florida Reef Tract  \\
        \textbf{GoM}   & Gulf of Mexico \\
        \textbf{HYCOM} & HYbrid Coordinate Ocean Model \\
        \textbf{LC}    & Loop Current \\
        \textbf{NOAA}  & National Oceanographic and Atmospheric Administration \\
        \textbf{PoM}   & Port of Miami \\
        \textbf{PMDDP} & PortMiami Deep Dredge Project \\
        \textbf{RS}    & Radiation stress \\
        \textbf{SCTLD} & Stony coral tissue loss disease \\
        \textbf{SLIM}  & Second-generation Louvain-la-Neuve Ice-ocean Model\\
        \textbf{SWAN}  & Simulating WAves Nearshore \\
        \textbf{WFS}   & West Florida Shelf
    \end{tabular}
\endgroup
